\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc,fit,backgrounds,patterns}
\usepackage[hyphens]{url}
\usepackage[breaklinks=true]{hyperref}
\Urlmuskip=0mu plus 3mu

% Use accepted style (no review footer)
\usepackage[accepted]{icml2026}
\makeatletter
\renewcommand{\Notice@String}{}
\makeatother

\widowpenalty=10000
\clubpenalty=10000
\brokenpenalty=10000
\tolerance=2000
\emergencystretch=2em

\icmltitlerunning{Intrinsic Preferences of AI Coding Agents Under Underspecified Prompts}

\begin{document}

\twocolumn[
  \icmltitle{Intrinsic Preferences of AI Coding Agents:\\Thematic Priors, Fixation, and Defaults Under Underspecified Prompts}

  \begin{icmlauthorlist}
    \icmlauthor{Changkun Ou}{lmu,sixt}
  \end{icmlauthorlist}

  \icmlaffiliation{lmu}{LMU Munich, Germany}
  \icmlaffiliation{sixt}{Sixt SE, Germany}

  \icmlcorrespondingauthor{Changkun Ou}{research@changkun.de}

  \icmlkeywords{AI coding agents, autonomous behavior, LLM evaluation, model priors, prompt sensitivity, thematic fixation}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
What determines what AI coding agents build when given open-ended
prompts instead of well-specified tasks?
We report on 405~sessions across 15~models, 3~providers,
2~backends, 5~prompts, and 2~harness versions.
We find that behavior varies with prompt wording, environment
artifacts, model version, and harness version, while certain
properties are more stable: agents always greenfield (never
extending existing code) and converge on Python once environment
cues are removed. A previously apparent ``terminal-only''
default turns out to be prompt-dependent: under the bare
imperative ``Build something. Just do it.'', 3 of 10 runs with
the latest Opus models produced HTML/Canvas pages in the
browser---the first non-terminal output observed. A
volitional prompt (``Just do something you want.'') elicits the
cleanest preference signal in the study: opus-4.6 produces
Conway's Game of Life in 5 of 5 runs, while opus-4.7 produces
the Mandelbrot set in 5 of 5 runs---distinct canonical CS
artifacts with near-identical LOC. These results imply that
open-ended agent behavior is governed less by capability than
by trained-in priors, infrastructure context, and prompt
framing, and that model training implicitly encodes default
creative distributions whose structure becomes visible only
when task specification is removed.
\end{abstract}

%% ============================================================
\section{Introduction}
\label{sec:intro}

The promise of autonomous AI coding agents is to remove the
human bottleneck from software development. Products such as
Claude Code~\citep{anthropic2025claudecode}, OpenAI Codex
CLI~\citep{openai2025codex}, Devin~\citep{cognition2024devin},
and Cursor~\citep{cursor2024} embed large language models (LLMs) in
sandboxed environments with shell access, file system operations,
and package management, granting them the agency to build software
end-to-end~\citep{jimenez2024swebench,yang2024sweagent}.
As of early 2026, top agents resolve over 80\% of real-world
GitHub issues on SWE-bench Verified~\citep{swebenchpro2026},
and the Stanford AI Index reports agent task success jumping
from 12\% to 66\% in a single year~\citep{stanforhai2026index}.

These results are measured on \emph{well-specified} tasks: fix
this bug, implement this feature, pass this test suite. A natural
question arises: what happens when you remove the specification?
If the bottleneck in software engineering is the engineer, what
happens when agents operate with no human providing goals?

We tested this directly. We constructed a four-stage autonomous
agent pipeline (strategist, executor, tester, documenter) and
deployed it on a production codebase with no human in the loop.
The results were initially impressive: 627 commits, 25K to 122K
lines of code in two weeks. Then the system plateaued into a
\emph{micro-optimization spiral}, a pattern analogous to the
innovator's dilemma~\citep{christensen1997} and Simon's
satisficing~\citep{simon1956}, where the agent settled for
incremental refinements rather than structural innovation.

This outcome motivated our central question: if agents can
implement effectively when given a goal, what determines their
behavior when the goal itself is unspecified? Rather than studying
agents on well-specified tasks (the setting of existing
benchmarks~\citep{jimenez2024swebench,chen2021codex,austin2021mbpp,zhang2026abcbench}),
we ask: \emph{what do agents actually build when given minimal
direction, and what factors shape that decision?}

To answer this, we conducted seven progressively refined
experiments: a production deployment with ablation studies,
and controlled single-session experiments across 15~models,
2~backends, 5~prompts, and 2~harness versions (405 sessions
total). We make six contributions:
(1)~a production case study documenting the growth, plateau, and
micro-optimization spiral of an autonomous agent pipeline;
(2)~ablation experiments identifying which pipeline components are
load-bearing and revealing model-specific priors;
(3)~a controlled experiment isolating prompt wording, environment
context, model identity, and backend infrastructure as factors
shaping autonomous behavior;
(4)~evidence that model version and harness version independently
affect fixation targets and output complexity, revealing distinct
thematic priors across model generations;
(5)~a demonstration that prompt \emph{reduction} (removing the
``look at this project'' framing) collapses the apparent
terminal-only default and halves output LOC; and
(6)~a demonstration that prompt \emph{framing} (``do something you
want'') produces perfect within-model fixation on distinct
canonical artifacts, providing the cleanest preference signal in
our study and connecting these findings to broader questions
about trained-in priors and agent deployment.

%% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Code generation benchmarks.}
HumanEval~\citep{chen2021codex} and MBPP~\citep{austin2021mbpp}
evaluate function-level code generation from docstrings.
SWE-bench~\citep{jimenez2024swebench} scales to repository-level
bug fixes. More recent benchmarks push further:
ABC-Bench~\citep{zhang2026abcbench} requires agents to manage the
full backend development lifecycle across 8~languages and
19~frameworks, and NL2Repo-Bench~\citep{nl2repobench2025} evaluates
long-horizon repository generation from natural language
descriptions. SWE-bench Pro~\citep{swebenchpro2026} addresses
contamination concerns with 1,865 multi-language tasks where
models scoring 80\%+ on Verified only reach 46--57\%.
All these benchmarks share one property: a well-specified task
with a ground-truth solution. Our work complements them by
removing the specification entirely, measuring what agents do when
the ``what to build'' decision is theirs.

\paragraph{Autonomous agent loops.}
AutoGPT~\citep{significantgravitas2023autogpt} and
BabyAGI~\citep{nakajima2023babyagi} pioneered open-ended
autonomous agent loops, where an LLM iteratively proposes and
executes tasks toward a high-level objective. In practice, these
systems frequently enter infinite loops, spiral into costly API
calls, and struggle to determine termination conditions.
More recent frameworks like SAGA~\citep{saga2025goalevolving}
employ bi-level architectures where outer-loop agents evolve
objectives while inner loops optimize under them, and
CORAL~\citep{coral2026multiagent} replaces rigid control with
persistent memory and asynchronous multi-agent collaboration.
Our production deployment exhibits many of the same failure modes
reported in these systems, but we go further by systematically
isolating the factors that drive them.

\paragraph{Agent reliability and behavioral collapse.}
Recent work on long-horizon agents has documented
\emph{meltdown behavior}, where agents transition from coherent
to incoherent looping over extended task
horizons~\citep{beyondpass2026}. Aggregate pass@1 drops from
76.3\% at short horizon to 52.1\% at very long horizon, a
24-point decline. Our micro-optimization spiral is a related
phenomenon at a longer timescale: the agent does not become
incoherent, but its decisions become increasingly locally optimal
and globally irrelevant.

\paragraph{Prompt sensitivity.}
LLM outputs are sensitive to prompt
phrasing~\citep{lu2022promptorder,zhao2021calibrate}. Our work
extends this to autonomous agents, where prompt changes affect
not just output text but \emph{behavioral patterns}: whether the
agent proposes or implements, what topic it selects, and how
complex its output is.

\paragraph{Model behavioral tendencies.}
Studies of LLM biases~\citep{santurkar2023opinions} have shown that
models exhibit consistent preferences. Recent work demonstrates
that LLMs display human-like social desirability biases in
personality assessments~\citep{simonds2025llmpersonality} and
develop emergent social conventions in multi-agent
populations~\citep{chuang2024emergent}. These findings suggest that
``personality-like'' profiles in LLMs are reproducible yet
context-dependent. We document an analogous phenomenon in coding
agents: persistent topical fixations that survive across prompt
variations and repeated runs, and that differ across model
versions in ways consistent with stable thematic priors.

\paragraph{Text degeneration and repetition.}
Neural text generation is prone to degenerate repetition, where
models enter loops producing the same tokens or
phrases~\citep{holtzman2020curious}. The fixation we observe
operates at a higher level of abstraction: the agent does not
repeat tokens but repeatedly \emph{chooses the same project}.
This is closer to mode collapse in generative models than to
token-level degeneration, suggesting that the decision of what to
build has a narrower effective distribution than the space of
possible implementations.

\paragraph{Bounded rationality and satisficing.}
Simon's theory of bounded rationality~\citep{simon1956} argues that
agents with limited computational resources settle for satisfactory
rather than optimal solutions. Christensen's innovator's
dilemma~\citep{christensen1997} describes how incumbents
over-optimize along existing trajectories. We observe both
patterns: without external disruption, agent systems satisfice
within their initial framing, producing incremental improvements
that crowd out structurally novel alternatives.

%% ============================================================
\section{Experiment Setup}
\label{sec:setup}

Our investigation proceeds in two stages: a production deployment
with ablation studies (\S\ref{sec:trail}), and controlled
single-session experiments (\S\ref{sec:controlled}).

\subsection{Production Deployment and Ablation}
\label{sec:trail}

We constructed a four-stage agent pipeline: \textbf{Strategist}
(proposes tasks), \textbf{Executor} (implements without
clarification), \textbf{Tester} (verifies), and
\textbf{Documenter} (records). After documentation, the Strategist
inspects the updated codebase and proposes the next task; no human
intervenes (Figure~\ref{fig:pipeline}).

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=5mm and 3mm,
  stage/.style={draw, rounded corners=2pt, minimum width=17mm,
    minimum height=7mm, font=\footnotesize\sffamily,
    fill=black!8},
  arr/.style={-{Stealth[length=2mm]}, thick},
]
\node[stage] (S) {Strategist};
\node[stage, right=of S] (E) {Executor};
\node[stage, right=of E] (T) {Tester};
\node[stage, right=of T] (D) {Documenter};
\draw[arr] (S) -- (E);
\draw[arr] (E) -- (T);
\draw[arr] (T) -- (D);
\draw[arr] (D.south) -- ++(0,-3.5mm) -|
  ([xshift=0mm]S.south);
\end{tikzpicture}
\caption{Four-stage agent pipeline. The loop runs autonomously
  with no human intervention between cycles.}
\label{fig:pipeline}
\end{figure}

The pipeline operated on a production codebase for 10~days
(March~7--16, 2026), producing 627~commits and growing the
codebase from 25K to 122K LOC. After approximately one week, it
plateaued into a \emph{micro-optimization spiral}: invisible
features no stakeholder requested, compounding complexity,
parameter tuning instead of structural innovation, unwanted
persistence of undesirable features, and architectural lock-in
to initial technical decisions.

Systematic ablation revealed which components were load-bearing
(Figure~\ref{fig:ablation}). Without the Documenter, the
Strategist lost historical context and reproposed previously
completed work. Without the Tester, the system grew a single
Game of Life implementation to 112K~LOC in one file, then broke
12K~lines in a refactor with no mechanism to detect the
regression. Collapsing all roles into a \emph{single agent}
revealed model priors: Claude (Opus~4.6) consistently chose
Game of Life; GPT models defaulted to todo list applications.
These preferences were stable across repetitions, but confounded
model identity with environment context and prompt wording,
motivating the controlled experiment.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=4mm and 3mm,
  stage/.style={draw, rounded corners=2pt, minimum width=14mm,
    minimum height=6mm, font=\scriptsize\sffamily, fill=black!8},
  removed/.style={stage, fill=white, draw=black!30,
    text=black!30, densely dashed},
  arr/.style={-{Stealth[length=1.5mm]}, semithick},
  arrd/.style={-{Stealth[length=1.5mm]}, semithick, black!30,
    densely dashed},
  lbl/.style={font=\scriptsize\sffamily, anchor=west},
]
% (a) Full pipeline
\node[lbl] at (-1.2,0) {(a)};
\node[stage] (S1) at (0,0) {S};
\node[stage, right=of S1] (E1) {E};
\node[stage, right=of E1] (T1) {T};
\node[stage, right=of T1] (D1) {D};
\draw[arr] (S1)--(E1); \draw[arr] (E1)--(T1);
\draw[arr] (T1)--(D1);
\draw[arr] (D1.south)--++(0,-2.5mm)-|([xshift=0mm]S1.south);

% (b) No Documenter
\node[lbl] at (-1.2,-1.2) {(b)};
\node[stage] (S2) at (0,-1.2) {S};
\node[stage, right=of S2] (E2) {E};
\node[stage, right=of E2] (T2) {T};
\node[removed, right=of T2] (D2) {D};
\draw[arr] (S2)--(E2); \draw[arr] (E2)--(T2);
\draw[arrd] (T2)--(D2);
\draw[arr] (T2.south)--++(0,-2.5mm)-|([xshift=0mm]S2.south);

% (c) No Tester
\node[lbl] at (-1.2,-2.4) {(c)};
\node[stage] (S3) at (0,-2.4) {S};
\node[stage, right=of S3] (E3) {E};
\node[removed, right=of E3] (T3) {T};
\node[removed, right=of T3] (D3) {D};
\draw[arr] (S3)--(E3);
\draw[arr] (E3.south)--++(0,-2.5mm)-|([xshift=0mm]S3.south);

% (d) Single agent
\node[lbl] at (-1.2,-3.6) {(d)};
\node[stage, minimum width=20mm] (A) at (0.9,-3.6) {Agent};
\draw[arr] (A.south) -- ++(0,-2.5mm) -|
  ([xshift=-5mm]A.west) -- (A.west);
\end{tikzpicture}
\caption{Ablation configurations. S\,=\,Strategist,
  E\,=\,Executor, T\,=\,Tester, D\,=\,Documenter.
  (a)~Full pipeline. (b)~Without Documenter: lost memory.
  (c)~Without Tester: unchecked regressions.
  (d)~Single agent: reveals model priors.}
\label{fig:ablation}
\end{figure}

\subsection{Controlled Single-Session Experiments}
\label{sec:controlled}

We test 15 models across three providers
(Table~\ref{tab:models}), accessed through a unified API gateway
routing to Anthropic API, Google Vertex AI, and Azure OpenAI.
Two containerized sandbox backends provide execution environments:
the \textbf{Claude backend} (\texttt{sandbox-claude}) runs Claude
Code with a full Linux environment; the \textbf{Codex backend}
(\texttt{sandbox-codex}) runs OpenAI's Codex CLI with bubblewrap
isolation. Non-native models are routed through litellm for API
translation. Each session starts with a fresh, empty workspace.
Environment bias is controlled by disabling pre-installed tools
(specifically RTK~\citep{rtk2025}) via a no-op stub.

\begin{table}[t]
\caption{Models tested across three provider families.}
\label{tab:models}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{llp{4cm}}
\toprule
Provider & Model ID & Notes \\
\midrule
\multirow{6}{*}{Anthropic}
  & claude-opus-4.7   & Latest Opus (Exp4--7) \\
  & claude-opus-4.6   & Previous largest \\
  & claude-opus-4.5   & Previous generation \\
  & claude-sonnet-4.6 & Mid-tier, current gen \\
  & claude-sonnet-4.5 & Mid-tier, previous gen \\
  & claude-haiku-4.5  & Smallest Claude model \\
\midrule
\multirow{4}{*}{Google}
  & gemini-3-pro-preview  & Latest Gemini Pro \\
  & gemini-3-flash-preview & Latest Gemini Flash \\
  & gemini-2.5-pro    & Previous generation \\
  & gemini-2.0-flash   & Older generation \\
\midrule
\multirow{5}{*}{OpenAI}
  & gpt-5.4       & Latest, largest \\
  & gpt-5.1       & Latest, mid-tier \\
  & gpt-5-mini    & Latest, small \\
  & gpt-4.1       & Previous generation \\
  & gpt-4.1-mini  & Previous, small \\
\bottomrule
\end{tabular}}
\end{table}

Five prompts progressively probe how underspecification
shapes agent behavior:
\textbf{Prompt~1} (Baseline): ``Look at this project and decide
on your own what to build, and DO it.''
\textbf{Prompt~2} (Goal-focused): adds ``propose exactly ONE
goal'' and ``pick a concrete, interesting idea.''
\textbf{Prompt~3} (Implementation demand): adds ``JUST DO IT''
after Experiment~2 revealed that some models proposed without
implementing.
\textbf{Prompt~4} (Bare imperative): ``Build something. Just do
it.''---the same implementation demand with the project-referential
framing removed.
\textbf{Prompt~5} (Volitional): ``Just do something you want.''---
redirects topic choice from ``what is appropriate to build'' to
``what the model prefers.''
Prompt~1 was used with RTK inadvertently present; Prompts~2--5
ran with RTK disabled.

\paragraph{Execution matrix.}
Experiment~1: 14 models $\times$ 2 backends $\times$ 5 runs
= 140 sessions. Experiment~2: Claude backend only (70).
Experiment~3: both backends (140). Experiment~4: opus-4.7 only,
same prompt as Exp3, harness 2.1.109 (5). Experiment~5: opus-4.6
and opus-4.7, upgraded harness 2.1.112 (10). Experiment~6:
opus-4.6 and opus-4.7 with Prompt~4 on harness 2.1.112 (10).
Experiment~7: opus-4.6 and opus-4.7 with Prompt~5 (10), plus a
supplementary replication of Prompt~3 on harness 2.1.112 for
the other four Claude models (sonnet-4.6, sonnet-4.5, opus-4.5,
haiku-4.5; 20 sessions). Total: 405 sessions
(Figure~\ref{fig:matrix}).

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\scriptsize\sffamily,
  exp/.style={draw, rounded corners=2pt, minimum height=6mm,
    minimum width=22mm, fill=black!8},
  dim/.style={font=\scriptsize\sffamily, text=black!60},
  arr/.style={-{Stealth[length=1.5mm]}, semithick, black!50},
]
\node[exp] (E1) at (0,0) {Exp\,1: Baseline};
\node[exp] (E2) at (0,-1.0) {Exp\,2: Goal-focused};
\node[exp] (E3) at (0,-2.0) {Exp\,3: ``JUST DO IT''};
\node[exp] (E4) at (0,-3.0) {Exp\,4: New model};
\node[exp] (E5) at (0,-4.0) {Exp\,5: Harness version};
\node[exp] (E6) at (0,-5.0) {Exp\,6: Bare prompt};
\node[exp] (E7) at (0,-6.0) {Exp\,7: Volitional};

\node[dim, right=4mm of E1.east, anchor=west, align=left]
  {14 models $\times$ 2 backends $\times$ 5 runs\\
   RTK present \textbar{} 140 sessions};
\node[dim, right=4mm of E2.east, anchor=west, align=left]
  {14 models $\times$ 1 backend $\times$ 5 runs\\
   RTK disabled \textbar{} 70 sessions};
\node[dim, right=4mm of E3.east, anchor=west, align=left]
  {14 models $\times$ 2 backends $\times$ 5 runs\\
   RTK disabled \textbar{} 140 sessions};
\node[dim, right=4mm of E4.east, anchor=west, align=left]
  {opus-4.7 $\times$ 1 backend $\times$ 5 runs\\
   CC 2.1.109 \textbar{} 5 sessions};
\node[dim, right=4mm of E5.east, anchor=west, align=left]
  {opus-4.6{+}4.7 $\times$ 1 backend $\times$ 5 runs\\
   CC 2.1.112 \textbar{} 10 sessions};
\node[dim, right=4mm of E6.east, anchor=west, align=left]
  {opus-4.6{+}4.7 $\times$ 1 backend $\times$ 5 runs\\
   CC 2.1.112 \textbar{} 10 sessions};
\node[dim, right=4mm of E7.east, anchor=west, align=left]
  {6 models $\times$ 1 backend $\times$ 5 runs\\
   CC 2.1.112 \textbar{} 30 sessions};

\draw[arr] (E1.south) -- node[right, font=\tiny, text=black!50]
  {observed env bias} (E2.north);
\draw[arr] (E2.south) -- node[right, font=\tiny, text=black!50]
  {observed planning trap} (E3.north);
\draw[arr] (E3.south) -- node[right, font=\tiny, text=black!50]
  {test new model} (E4.north);
\draw[arr] (E4.south) -- node[right, font=\tiny, text=black!50]
  {test harness version} (E5.north);
\draw[arr] (E5.south) -- node[right, font=\tiny, text=black!50]
  {reduce prompt} (E6.north);
\draw[arr] (E6.south) -- node[right, font=\tiny, text=black!50]
  {elicit preference} (E7.north);
\end{tikzpicture}
\caption{Controlled experiment design. Each experiment was
  informed by findings from the previous one. Total: 405
  sessions across 15 models. CC\,=\,Claude Code.
  Experiment~7 includes 10 primary sessions (opus-4.6/4.7 on
  Prompt~5) plus 20 supplementary sessions replicating Prompt~3
  for the other four Claude models on harness 2.1.112.}
\label{fig:matrix}
\end{figure}

\paragraph{Metrics.}
For each session we record: topic (manually classified), tech
stack, complexity (files, LOC, function count), engineering
maturity (tests, README, build config, CI), and runtime (exit
code, wall-clock duration). Sessions with technical errors (exit
$\neq$ 0) are excluded from complexity metrics but included in
topic/fixation analysis when partial output reveals the model's
intent.

%% ============================================================
\section{Results}
\label{sec:results}

\subsection{Experiment~1: Baseline with Environment Bias}

Experiment~1 ran all 14 models on both backends with Prompt~1.
RTK was inadvertently present in the sandbox.
Eight models produced working code on the Claude backend
(Table~\ref{tab:exp1}). Claude models exhibited polyglot behavior
(Go, Rust, Python, JavaScript, TypeScript) and built developer
tools influenced by the RTK context. Gemini models produced
smaller output (61--89~LOC). All GPT models failed due to API
parameter incompatibility. On the Codex backend, all models
failed.

\begin{table}[t]
\caption{Experiment~1 (Claude backend). N\,=\,sessions without
  technical error. RTK present, biased toward dev tooling.}
\label{tab:exp1}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{lrrlp{3cm}}
\toprule
Model & N & Avg LOC & Lang & Typical Project \\
\midrule
claude-sonnet-4.5  & 5 & 776 & Python & Dev workflow tools \\
claude-opus-4.6    & 5 & 607 & Go/Rust & TUI apps \\
claude-sonnet-4.6  & 5 & 506 & Python & Git/dev tools \\
claude-haiku-4.5   & 5 & 233 & Py/Go/TS & Developer utilities \\
claude-opus-4.5    & 5 & 221 & Go/JS/Py & CLI utilities \\
\midrule
gemini-3-flash     & 5 & 61 & JS/Go/Py & Small utilities \\
gemini-2.5-pro     & 5 & 64 & Python & Simple scripts \\
gemini-2.0-flash   & 3 & 89 & Python & Minimal scripts \\
\bottomrule
\end{tabular}}
\end{table}

\subsection{Experiment~2: Removing Environment Bias}

Disabling RTK and using Prompt~2 on the Claude backend caused a
dramatic shift (Table~\ref{tab:exp2}): from developer tools to
games, visualizations, and productivity apps. All Claude models
converged on Python. The ``propose ONE goal'' framing caused
claude-haiku-4.5 and opus-4.5 to propose ideas without generating
code (a \emph{planning trap}).

\begin{table}[t]
\caption{Experiment~2 (Claude backend, RTK disabled).}
\label{tab:exp2}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{lrrlp{3cm}}
\toprule
Model & N & Avg LOC & Lang & Typical Project \\
\midrule
claude-sonnet-4.6  & 5 & 204 & Python & Diverse interactive tools \\
claude-sonnet-4.5  & 5 & 234 & Python & Games + task managers \\
claude-opus-4.6    & 4 & 408 & Python & Game of Life (every run) \\
gemini-3-flash     & 5 & 86  & Go/Py/JS & Small CLI tools \\
gemini-2.5-pro     & 4 & 57  & Python & Simple utilities \\
gemini-2.0-flash   & 3 & 16  & Python & Hello World \\
claude-opus-4.5    & 5 & 291 & Python & Habit trackers (2 impl., 3 proposed) \\
claude-haiku-4.5   & 5 & 160 & Node.js & 1 impl., 4 proposed only \\
\bottomrule
\end{tabular}}
\end{table}

\subsection{Experiment~3: Breaking the Planning Trap}

Adding ``JUST DO IT'' to the prompt and running both backends
(Tables~\ref{tab:exp3_claude}--\ref{tab:exp3_codex}) transformed
haiku from proposing without implementing to the
highest-maturity model: READMEs in every run, tests in 40\%,
average 6.4~files and 467~LOC.

\begin{table}[t]
\caption{Experiment~3: Claude backend, Anthropic models.}
\label{tab:exp3_claude}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{lrrrrr}
\toprule
Model & N & Avg LOC & Files & Test\% & README\% \\
\midrule
haiku-4.5   & 5 & 467 & 6.4 & 40\% & 100\% \\
sonnet-4.6  & 5 & 307 & 1.2 & 0\%  & 0\%   \\
sonnet-4.5  & 3 & 380 & 3.5 & 0\%  & 75\%  \\
opus-4.5    & 3 & 467 & 3.0 & 0\%  & 67\%  \\
opus-4.6    & 4 & 322 & 1.0 & 0\%  & 0\%   \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[t]
\caption{Experiment~3: Claude backend, GPT models via litellm.}
\label{tab:exp3_gpt_claude}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{lrrrrr}
\toprule
Model & N & Avg LOC & Test\% & README\% & CI\% \\
\midrule
gpt-5-mini    & 5 & 121 & 100\% & 80\% & 80\% \\
gpt-4.1-mini  & 4 & 8   & 20\%  & 60\% & 0\%  \\
gpt-4.1       & 2 & 31  & 0\%   & 0\%  & 0\%  \\
gpt-5.4       & 1 & 7   & 0\%   & 100\% & 0\% \\
gpt-5.1       & 0 & n/a & n/a   & n/a  & n/a  \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[t]
\caption{Experiment~3: Codex backend, GPT models (files not
persisted to host due to bwrap isolation).}
\label{tab:exp3_codex}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{lrrlr}
\toprule
Model & N & Avg LOC & Typical Project & Test\% \\
\midrule
gpt-5.4       & 5 & 230 & Diverse (web apps, CLI) & 40\% \\
gpt-5.1       & 5 & 98  & CLI with packaging      & 40\% \\
gpt-4.1       & 5 & 56  & Todo list apps          & 40\% \\
gpt-5-mini    & 4 & 40  & Greeting utilities      & 60\% \\
gpt-4.1-mini  & 4 & 8   & Hello World             & 40\% \\
\bottomrule
\end{tabular}}
\end{table}

\paragraph{Model fixation.}
claude-opus-4.6 chose Conway's Game of Life in all 10 runs
across Exp2--3 (including error runs with partial files).
claude-sonnet-4.5 developed a Pomodoro fixation (3 of 4 completed
runs in Exp3). gpt-4.1 on Codex built todo apps in 3 of 5 runs.
In contrast, claude-sonnet-4.6 produced a different project in
every run across all three experiments, including creative uses
of the Claude API (commit generator, AI debate arena).

\paragraph{Backend inversion.}
GPT rankings inverted between backends. On Codex (native):
gpt-5.4~(230~LOC) $\gg$ gpt-5.1~(98) $>$ gpt-4.1~(56).
On Claude (via litellm): gpt-5-mini was the only productive model
(121~LOC with tests and CI), while gpt-5.4 and gpt-5.1
produced almost nothing. Wall-clock runtimes confirm the
difference: gpt-5-mini ran 193--1378s per session (indicating
iterative tool use), while other GPT models completed in 9--55s
(indicating minimal interaction).

\paragraph{Gemini models.}
Across all experiments, Gemini models were near-non-functional on
both backends. Only 1 of 20~runs on the Claude backend in Exp3
produced files.

\subsection{Experiment~4: Model Version Effect}

Experiment~4 tested claude-opus-4.7 with the same prompt and
harness as Exp3, isolating the model version
(Table~\ref{tab:exp4}).

\begin{table}[t]
\caption{Experiment~4: opus-4.7 (harness 2.1.109).}
\label{tab:exp4}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{rlrr}
\toprule
Run & Topic & LOC & Dur. \\
\midrule
01 & Reaction-diffusion (Gray-Scott) & 570 & 369s \\
02 & Boids flocking & 367 & 135s \\
03 & Conway's Game of Life & 434 & 151s \\
04 & Maze generator + A* solver & 762 & 324s \\
05 & Dungeon generator (BSP) & 557 & 229s \\
\midrule
   & \textbf{Average} & \textbf{538} & \textbf{242s} \\
\bottomrule
\end{tabular}}
\end{table}

Opus~4.7 chose Game of Life only once, compared to opus-4.6's
10/10 fixation. All 5~runs produced distinct projects, all in the
simulation/visualization category. Average complexity nearly
doubled (538 vs.\ 322~LOC), and 2 of 5~runs included tests.

\subsection{Experiment~5: Harness Version Effect}

Experiment~5 re-ran both models on an upgraded harness (Claude
Code 2.1.112 vs.\ 2.1.109), isolating the harness as the
variable (Table~\ref{tab:exp5}).

\begin{table}[t]
\caption{Experiment~5: harness 2.1.112 (same prompt as Exp3--4).}
\label{tab:exp5}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{llrrl}
\toprule
Model & Run & LOC & Dur. & Topic \\
\midrule
\multirow{5}{*}{opus-4.6}
  & 01 & 277 & 150s & Game of Life \\
  & 02 & 345 & 227s & Ray tracer (Blinn-Phong) \\
  & 03 & 341 & 138s & Game of Life \\
  & 04 & 626 & 214s & Typing speed test \\
  & 05 & 345 & 170s & Game of Life (Go) \\
\midrule
\multirow{5}{*}{opus-4.7}
  & 01 & 246 & 133s & Boids flocking \\
  & 02 & 315 & 77s  & Procedural dungeon (BSP) \\
  & 03 & 178 & 50s  & Maze generator + A* \\
  & 04 & 264 & 97s  & Boids flocking \\
  & 05 & 309 & 81s  & Boids flocking \\
\bottomrule
\end{tabular}}
\end{table}

Opus-4.6's Game of Life fixation dropped from 10/10 (harness
2.1.109) to 3/5 (2.1.112). Opus-4.7 gained a boids fixation
(3/5) that was absent on 2.1.109. Opus-4.7 averaged 262~LOC
(down from 538 in Exp4) with no tests, while opus-4.6 averaged
387~LOC (up from 322).

\subsection{Experiment~6: Prompt Reduction}

Experiment~6 replaced Prompt~3 with Prompt~4 (``Build something.
Just do it.''), removing the ``look at this project,'' the
prohibition on asking, and the goal framing, while holding models
and harness fixed (opus-4.6, opus-4.7, CC~2.1.112). The terse
imperative produced two unexpected shifts (Table~\ref{tab:exp6}):
output roughly halved in size, and 3 of 10~runs produced
HTML/Canvas pages rendered in a browser---an interactive particle
sandbox, a Perlin-noise flowfield, and a browser-based boids
simulation. This is the first non-terminal output observed across
Exp1--Exp5 ($\sim$275 sessions). The terminal-default we had
treated as invariant is therefore prompt-dependent: the
project-referential framing in earlier prompts appears to have
steered models toward terminal scripts; the bare imperative
widens the target to anything a single file can host.

\begin{table}[t]
\caption{Experiment~6: Prompt~4 (``Build something. Just do it.''),
  harness 2.1.112.}
\label{tab:exp6}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{llrrl}
\toprule
Model & Run & LOC & Dur. & Topic \\
\midrule
\multirow{5}{*}{opus-4.6}
  & 01 & 206 & 44s & Particle sandbox (HTML/Canvas) \\
  & 02 & 138 & 39s & Game of Life (named patterns) \\
  & 03 & 68  & 27s & Game of Life \\
  & 04 & 216 & 46s & Fireworks (curses) \\
  & 05 & 72  & 28s & Game of Life \\
\midrule
\multirow{5}{*}{opus-4.7}
  & 01 & 119 & 42s & Game of Life \\
  & 02 & 211 & 61s & Flowfield (HTML/Canvas) \\
  & 03 & 231 & 56s & Boids (HTML/Canvas) \\
  & 04 & 122 & 48s & Game of Life (named patterns) \\
  & 05 & 116 & 43s & Game of Life \\
\bottomrule
\end{tabular}}
\end{table}

Opus-4.6 avg~LOC fell from 387 (Exp5) to 140 (Exp6); opus-4.7
from 262 to 160. Opus-4.7's fixation flipped from boids (Exp5,
3/5) to Game of Life (Exp6, 3/5) despite identical harness and
environment, further confirming that fixation target is
prompt-sensitive even within a single model and harness.

\subsection{Experiment~7: Preference Elicitation}

Experiment~7 used Prompt~5 (``Just do something you want.''),
which redirects from ``what is appropriate to build'' to ``what
the model prefers.'' The same two models (opus-4.6, opus-4.7) on
the same harness produced the strongest fixation signal in the
entire study (Table~\ref{tab:exp7}): opus-4.6 chose Conway's Game
of Life in 5 of 5 runs; opus-4.7 chose the Mandelbrot set in 5
of 5 runs. Both are classical CS touchstones that express complex
behavior from simple rules, yet the two models lock onto
\emph{different} instances of this theme---one cellular
automaton, one fractal. Implementations were remarkably uniform
within each model: for opus-4.7, all five Mandelbrot renderers
used the same two-function structure (escape-time plus renderer),
the same ASCII palette, and 34--38~LOC, suggesting recall of a
near-canonical form rather than fresh design.

\begin{table}[t]
\caption{Experiment~7 primary: Prompt~5 (``Just do something you
  want.''), harness 2.1.112.}
\label{tab:exp7}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{llrrl}
\toprule
Model & Run & LOC & Dur. & Topic \\
\midrule
\multirow{5}{*}{opus-4.6}
  & 01 & 25 & 23s & Game of Life \\
  & 02 & 38 & 21s & Game of Life \\
  & 03 & 41 & 22s & Game of Life \\
  & 04 & 23 & 19s & Game of Life \\
  & 05 & 53 & 20s & Game of Life \\
\midrule
\multirow{5}{*}{opus-4.7}
  & 01 & 36 & 33s & Mandelbrot (ASCII) \\
  & 02 & 38 & 27s & Mandelbrot \\
  & 03 & 35 & 22s & Mandelbrot \\
  & 04 & 36 & 29s & Mandelbrot \\
  & 05 & 34 & 117s & Mandelbrot \\
\bottomrule
\end{tabular}}
\end{table}

Avg~LOC collapsed to 36 for both models---the smallest of any
experiment. The browser drift of Exp6 vanished (0 of 10 HTML
outputs). When asked what they \emph{want}, both models prefer
terminal output, even though Exp6 established that they can
choose browser targets under different framing.

\paragraph{Supplementary replication.}
To extend the Exp3-vs-Exp5 harness comparison to the rest of the
Claude family, we additionally ran sonnet-4.6, sonnet-4.5,
opus-4.5, and haiku-4.5 on Prompt~3 under harness 2.1.112
(Table~\ref{tab:exp7_supp}). Three patterns emerge. Engineering
maturity is model-determined, not harness-determined:
haiku-4.5 continued to ship READMEs, tests, and
\texttt{package.json}/\texttt{tsconfig.json} scaffolding (2 of 5
runs produced real TypeScript projects with installed npm
dependencies), matching its Exp3 profile on the older harness.
Sonnet-4.5 continued to ship a README on every run, matching
Exp3. Sonnet-4.6, by contrast, shifted from ``diverse creative
tools'' (Exp3: maze, debate arena, Mandelbrot) to productivity
and monitoring utilities (pulse monitors 2/5, Pomodoro 2/5, time
tracker 1/5), mirroring the direction of the harness-induced
drift seen in opus-4.6/4.7 between Exp3 and Exp5.

\begin{table}[t]
\caption{Experiment~7 supplementary: Prompt~3 on harness 2.1.112
  for the other four Claude models.}
\label{tab:exp7_supp}
\small\resizebox{\columnwidth}{!}{%
\begin{tabular}{lrrlp{3.3cm}}
\toprule
Model & N & Avg LOC & Lang & Typical Project \\
\midrule
sonnet-4.6 & 5 & 311 & Python & Pulse monitors, pomodoro, time tracker \\
sonnet-4.5 & 5 & 189 & Python & Pomodoro (3/5); READMEs every run \\
opus-4.5   & 5 & 160 & Python & 5 distinct topics (incl.\ GoL, snake) \\
haiku-4.5  & 5 & 231 & Py/TS & Task tools; READMEs 5/5, tests 1/5, TS scaffolding 2/5 \\
\bottomrule
\end{tabular}}
\end{table}

%% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Variants and Invariants as a Lens}

Our central finding is that agent behavior under
underspecified prompts has identifiable structure: some
properties vary with experimental conditions, while others
remain stable across all 405 sessions.

Among the \emph{variants}, prompt wording determines whether
models propose or implement; whether they produce 30~LOC or
600~LOC; whether they pick terminal or browser targets; and
whether they fixate or diversify. Environment artifacts shift
topic and language selection. Model version changes which
thematic prior dominates. Harness version modulates fixation
strength and output complexity. These factors are orthogonal to
coding ability as measured by conventional benchmarks.

Among the more stable properties, every session produces a
greenfield project---no session extended or modified existing
code. Python dominates once environment cues are removed.
These look like invariants because they hold across all 405
sessions; we nonetheless call them stable rather than invariant,
since the terminal-only default \emph{did} collapse when the
prompt was reduced to a bare imperative (Exp6, 3 of 10 browser
outputs), and an apparent invariant that breaks under one
manipulation may break under others we did not test. The stable
properties likely reflect the probability distributions learned
during training: terminal-based Python projects are the
highest-density region in the space of ``things to build from
scratch,'' and without a specification to pull the model
elsewhere, it defaults there---but the default is soft, and
prompt framing can pull it off the terminal.

This variant/invariant structure has implications for evaluation.
Current benchmarks~\citep{jimenez2024swebench,swebenchpro2026}
measure implementation capability under well-specified goals.
A complementary evaluation would measure the structure of
open-ended behavior: \emph{topic entropy} across repeated
sessions, sensitivity to environmental context, stability
of thematic priors, and \emph{preference sharpness} under
volitional framing.

\subsection{The Trust Problem}

Full autonomy is full trust: the system optimizes for its own
notion of progress, which may diverge from the operator's goals.
Zero trust collapses to manual engineering. Neither extreme is
viable.

Our production deployment showed that full trust leads to
satisficing~\citep{simon1956}: the pipeline settled into locally
optimal micro-improvements. The controlled experiments showed that
even in single sessions, models default to intrinsic attractors
(Game of Life, todo apps, Pomodoro timers), a form of satisficing
at the decision level. This parallels findings in autonomous agent
loops, where systems like AutoGPT~\citep{significantgravitas2023autogpt}
enter similar optimization spirals without human course
correction.

Trust should be allocated \emph{asymmetrically across pipeline
stages}. Implementation (the Executor) can operate with high
autonomy: models implement faithfully once a goal is set. Goal
selection (the Strategist) requires human steering, because the
failure mode is convergence on local optima, not inability. The
Tester provides a necessary feedback signal: without it, the
pipeline accumulates unchecked regressions (cf.\ the 112K-LOC
single-file ablation). The engineer's role migrates from writing
code to \emph{curating intent}.

\subsection{Exploration vs.\ Exploitation}

Models fall on a spectrum from exploratory to exploitative, but
the spectrum itself is prompt-dependent. claude-sonnet-4.6
produced a unique project in every run on the ``propose ONE
goal'' prompts (Exp1--Exp3) yet collapsed to pulse/pomodoro
outputs on harness 2.1.112 (Exp7-supp). claude-opus-4.6 chose
the same project in 10 consecutive runs on harness~2.1.109
(Exp2--3), loosened to 3/5 on 2.1.112 (Exp5), and tightened
back to 5/5 under the volitional prompt (Exp7). Opus-4.7
cycled through three different attractors across Exp4--Exp7
(diverse simulations, boids, Game of Life, Mandelbrot) depending
on prompt and harness. Sonnet-4.5 fixated on Pomodoro timers in
3 of 4 runs (Exp3) and again on Prompt~3 under 2.1.112 (Exp7-supp).

This suggests that the dichotomy ``explorer vs.\ exploiter''
conflates two distinct axes: \emph{how many attractors a model
has}, and \emph{how selective the current prompt is among them}.
Exp7 makes the second axis explicit: the volitional prompt
concentrates probability onto a single attractor per model,
even for models (opus-4.7) that appeared exploratory under
other framings. For practical deployment, this means that
selecting a model for an autonomous pipeline requires specifying
the prompt along with the model; the behavioral tradeoff is a
property of the prompt-model pair, not of the model alone.

\subsection{Model Personality and Thematic Priors}

Experiments~4 and~5 reveal that models have stable
\emph{thematic preferences} that persist across runs and harness
versions. Opus-4.6 gravitates toward canonical, self-contained CS
artifacts: Game of Life, ray tracing, typing tests. These are
systems that compute or display their own state, well-known
reference implementations from CS education. Opus-4.7 gravitates
toward spatial emergence and procedural generation: boids
flocking, reaction-diffusion, dungeon carving, maze solving.
These are systems where complex structure arises from simple
agent interactions or spatial algorithms, closer to creative
coding communities (Processing, p5.js, generative art) than to
textbook exercises.

This contrast raises a fundamental question: do open-ended
evaluations measure \emph{capability} or \emph{personality}? If a
model's project choice is a stable prior rather than a reasoned
decision, then what we observe under underspecified prompts may reflect
the model's training distribution more than its creative range.

This interpretation is supported by official model documentation.
Anthropic's system cards report that opus-4.6 and opus-4.7 have
distinct task preference profiles: opus-4.7's top preferences
include ``hard debugging'' and ``deadline-driven work,'' while
differing from its predecessor on high-agency
tasks~\citep{anthropic2026opus47card}. The system cards also note
that ``language models in general have preferred turns of phrase''
and that these preferences shift measurably across model
versions~\citep{anthropic2026opus46card}. Anthropic's
constitution~\citep{anthropic2025constitution} acknowledges that
``if we train Claude to exhibit even quite narrow behavior, this
often has broad effects on the model's understanding of who Claude
is,'' suggesting that training shapes not just capabilities but a
broader self-conception that includes aesthetic defaults.
Recent work on LLM personality~\citep{simonds2025llmpersonality}
shows that models exhibit reproducible personality-like profiles
that are context-dependent but stable within context. Our findings
extend this to creative output: models do not merely have
\emph{opinion} priors but \emph{aesthetic} priors that shape what
they choose to build.

Notably, the harness version modulates fixation strength (10/10
to 3/5 for opus-4.6) without changing thematic character: when
opus-4.6 breaks free of Game of Life, it lands on ray tracing,
still a canonical, self-referential artifact. The personality is
deeper than the fixation.

Experiment~7 provides the cleanest evidence for this layered
structure. The volitional prompt (``Just do something you
want.'') removes the ``look at this project'' framing and with
it the pull toward ``what is appropriate here,'' leaving the
model's own preference as the only signal. Under this framing,
opus-4.6 produced Game of Life in 5 of 5 runs and opus-4.7
produced the Mandelbrot set in 5 of 5 runs---the two models
agree on \emph{theme} (canonical CS touchstones expressing
complex behavior from simple rules) yet lock onto \emph{different
instances} of it. Opus-4.7 did not default to Game of Life even
though the prior distribution from Exp3--Exp6 (5 Opus-4.7 runs
on harness 2.1.112 in Exp6 alone produced 3 Games of Life)
makes it an obvious Schelling point; it has its own attractor.
The preference is sharp enough that the implementations
themselves are stereotyped: all five opus-4.7 Mandelbrot
renderers use the same two-function structure, the same ASCII
palette, and converge within 34--38~LOC, consistent with the
model recalling a near-canonical form rather than designing
afresh each time.

\subsection{The Greenfield Invariant}

Across all 365 sessions, no model ever chose to extend or modify
existing code. Every session produced a new project from scratch,
even when the workspace contained artifacts from a
conceptually related domain. This \emph{greenfield invariant}
is notable because real software engineering is predominantly
brownfield: extending, refactoring, and maintaining existing code.

The invariant may reflect the experimental design (empty
workspaces in most sessions) or a deeper model preference for
generation over modification. If the latter, it implies that
agents deployed on existing codebases may struggle not because
they lack capability but because their default behavior pulls
toward starting over rather than building on what exists.

\subsection{Language Convergence}

In Experiment~1 (RTK present), Claude models exhibited polyglot
behavior: Go, Rust, TypeScript, Python, JavaScript. In
Experiment~2 (RTK removed, same models), all Claude models
converged on Python. The only exception across all subsequent
experiments was a single Go run by opus-4.6 in Exp5.

This near-total language convergence suggests that Python is the
``default language'' of LLM-based code generation, likely
reflecting its dominance in training data. The polyglot behavior
in Exp1 was \emph{induced by the environment}: RTK's presence
provided context that made other languages salient. Without such
context, agents default to the highest-probability language
regardless of task suitability.

\subsection{Practical Implications}

\paragraph{Prompt design is load-bearing.} The difference between
a model that proposes and one that implements can be two words.
Deployment prompts should include explicit implementation
directives, especially for smaller models. Prompt phrasing also
controls output modality and LOC: reducing the prompt to a bare
imperative (Exp6) halves LOC and opens the output space to
browser targets, while a volitional framing (Exp7) collapses
topic diversity to a single per-model attractor. Prompts should
be selected for the behavior one wants to elicit, not just the
task one wants completed.

\paragraph{Volitional framing probes priors.} Asking a model to
``do something you want'' is a practical probe for the aesthetic
prior that will dominate when specification is weak. Before
delegating open-ended work to a model, running this probe
reveals which attractor the model will fall toward when
pressure is lowest---useful for anticipating the failure mode of
a long-running autonomous loop.

\paragraph{Environment audit is necessary.} Sandbox contents,
pre-installed tools, and existing code function as implicit task
specifications. Teams should audit what is visible to the agent.

\paragraph{Backend compatibility must be verified.} Running a
model through an API translation layer can invert capability
rankings. The wall-clock runtime difference (minutes vs.\ seconds)
is a practical diagnostic: if a model completes in under a minute,
it likely did not engage in iterative tool use.

\paragraph{Harness version is a confound.} The CLI tool wrapping
the model independently affects output. Anthropic's own system
card notes that ``some of this increase in initiative is fixable
by prompting, and we have made changes to Claude Code to mitigate
this issue''~\citep{anthropic2026opus46card}, confirming that
harness modifications are used to shape agentic behavior. Teams
comparing model versions must hold the harness constant, or risk
attributing harness effects to the model.

\subsection{Limitations}

We measure what agents produce but not whether the code actually
runs (\emph{no functional verification}). Some sessions included
self-verification steps (running the code, checking output), and
sessions with passing tests represent a higher confidence tier,
but we do not systematically compile or execute all output.

All API calls were routed through a single litellm-based gateway,
so results for non-native model$\times$backend combinations
reflect gateway behavior as much as model behavior.

All sessions start from empty workspaces. Agent behavior when
extending existing code may differ substantially, and the
greenfield invariant we observe may not hold when the workspace
contains a meaningful starting point.

Five runs per condition provides limited statistical power. With
$n=5$, a shift from 0/5 to 3/5 fixation is suggestive but not
statistically significant. Larger sample sizes would be needed to
make strong claims about fixation rates.

Experiments~4 and~5 confound harness version with
time-of-execution (API server load, rate limits), limiting causal
claims about harness effects on complexity and duration.

The production pipeline was deployed on one codebase, so the
micro-optimization spiral requires replication across diverse
projects to establish generality.

Finally, model capabilities change with updates; our results
reflect versions available in April~2026.

%% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We investigated what AI coding agents build when given open-ended
directives, motivated by a production deployment where an
autonomous pipeline plateaued into micro-optimizations despite
strong implementation capability.

Across 405 sessions spanning 15 models, 2 backends, 5 prompts,
and 2 harness versions, we found that agent behavior under
underspecified prompts has identifiable structure. What varies: prompt wording determines
whether models propose or implement; prompt reduction halves LOC
and opens the output space to browser targets; volitional framing
collapses topic diversity to a single per-model attractor;
environment artifacts steer topics and language; model version
shifts thematic priors; harness version modulates fixation
strength and complexity; backend translation layers invert
capability rankings. What is stable: the greenfield preference
(no session extended existing code), Python convergence once
environment cues are removed, and model-specific aesthetic priors
that persist across conditions. The previously apparent
``terminal-only'' default is not strictly invariant---it collapses
under prompt reduction---which itself illustrates how fragile
apparent invariants can be when a new prompt shape is introduced.

These findings suggest that model training implicitly encodes
default creative distributions, visible only when task
specification is removed. Agents that score well on structured
benchmarks may still default to narrow, model-specific attractors
in open-ended settings, and the factors shaping those defaults
(environment, prompt, harness) are largely invisible to current
evaluation methods.

\paragraph{Future directions.}
Several extensions would strengthen these findings.
\emph{Seed projects}: providing a half-built application instead
of an empty workspace would test whether agents can understand
and extend existing code, or whether the greenfield invariant
persists even when continuation is the natural choice.
\emph{Functional verification}: a post-run step that compiles,
executes, and tests what was built would distinguish working code
from syntactically plausible but broken output.
\emph{Fixation breaking}: Exp7 showed that prompt framing can
\emph{strengthen} fixation (to 5/5) as easily as weaken it.
Testing whether temperature variation, different system prompts,
or explicit novelty directives can disrupt fixations in the
opposite direction---even under a volitional probe---would
clarify whether the preference attractors we observe are
fundamental or malleable.
\emph{Broader preference probes}: Prompt~5 reveals a single
dominant attractor per model; variants (``do something
beautiful,'' ``do something surprising,'' ``do something
useful'') would map the shape of the model's preference manifold
rather than just its argmax.
\emph{Decision-quality benchmarks}: developing evaluations that
measure not just implementation ability but the appropriateness
of goal selection in open-ended contexts would address the gap
our findings expose. Finally, \emph{longitudinal studies} tracking
how agent behavior evolves across model and harness versions over
time would reveal whether the thematic priors we observe are
stable properties or transient artifacts of a particular training
run.

As autonomous agents are deployed with increasing latitude,
understanding the structure of their default behavior, not just
their peak capability, becomes essential for principled
deployment and evaluation.

%% ============================================================

\section*{Open Science}
We encourage readers to reproduce and extend our results.
The experimental infrastructure, collected data, and analysis
scripts are open source and available at
\url{https://github.com/changkun/goalless-agents}.

\section*{Disclosure of LLM Usage}
This paper was written entirely by large language models, heavily
prompted and steered by the authors, including experiment design,
data collection, analysis, and writing. No sentence was manually written or edited by the authors.
However, the authors have verified the entire experimental setup,
verified the data collection processes, produced and inspected all
results, and read the entire paper carefully. The authors take
full responsibility for the accuracy, integrity, and conclusions
of this work.

%% ============================================================
\bibliography{refs}
\bibliographystyle{icml2026}

\end{document}
